# Pan-cancer expression signature modules and immune subtype identification

```{r setup_pkgs}
library(ggalign)
```

Data was downloaded from
https://www.cell.com/immunity/fulltext/S1074-7613(18)30121-3?utm_campaign=STMJ_1522958526_SC&utm_channel=WEB&utm_source=WEB&dgcid=STMJ_1522958526_SC#fig1 and
provided in this repository under the `FigureS6` folder.

## Data Loading and Preparation
```{r}
immune_subtypes <- readxl::read_xlsx(
    "FigureS6/immune_subtypes.xlsx",
    na = c("", "NA")
)
nrow(immune_subtypes)

immune_scores <- data.table::fread(
    "FigureS6/Scores_160_Signatures.tsv.gz"
)
mat <- cor(t(immune_scores[, !(1:2)]), method = "spearman")
signatures <- immune_scores[[2L]]
subtypes <- c(
    "Lymphocyte", "Macrophage", "TGF-beta", "IFN-gamma",
    "Wound Healing"
)
signatures[match(
    c(
        "LIexpression_score", "CSF1_response", "TGFB_score_21050467",
        "Module3_IFN_score", "CHANG_CORE_SERUM_RESPONSE_UP"
    ),
    signatures
)] <- subtypes

rownames(mat) <- colnames(mat) <- signatures

subtype_scores <- immune_subtypes |>
    dplyr::select(
        all_of(c(
            "Immune Subtype",
            "Wound Healing", "Macrophage Regulation",
            "Lymphocyte Infiltration Signature Score",
            "IFN-gamma Response", "TGF-beta Response"
        ))
    ) |>
    dplyr::filter(!is.na(`Immune Subtype`))
names(subtype_scores)[match(
    c(
        "Lymphocyte Infiltration Signature Score",
        "Macrophage Regulation", "TGF-beta Response",
        "IFN-gamma Response", "Wound Healing"
    ),
    names(subtype_scores)
)] <- subtypes
```

Five representative gene expression signatures (linked with colored polygons)
were used to cluster TCGA tumor samples (rows), resulting in the identification
of six immune subtypes, labeled C1– C6. The distribution of signature scores
across the six subtypes is shown, with the dashed line indicating the median
score for each subtype.
```{r}
p <- stack_crossv(mat, sizes = c(0.1, 1, 1)) -
    scheme_theme(
        plot.margin = margin(),
        axis.title = element_text(size = 10, face = "bold"),
        axis.text = element_text(size = 10, colour = "black"),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 14, face = "bold"),
        strip.text.y = element_text(size = 14)
    ) -
    scheme_align(NULL) +
    ggheatmap() +

    # we will highlight the subtypes
    geom_vline(
        aes(xintercept = .data$x),
        data = function(x) {
            x <- dplyr::filter(x, .data$.column_names %in% subtypes) |>
                dplyr::select(.x) |>
                dplyr::distinct() |>
                dplyr::pull(.x)
            data.frame(x = c(x - 0.5, x + 0.5))
        },
        linewidth = 0.1
    ) +
    theme_no_axes() +
    no_expansion() +
    labs(fill = "Spearman\ncorrelation") +

    ## add top dendrogram
    anno_top(size = 0.1) -
    scheme_align(free_spaces = "l") +
    align_dendro(method = "ward.D2") +
    no_expansion("y") +
    theme_no_axes("y") +

    ## add left dendrogram
    anno_left() -
    scheme_align(free_spaces = "b") +
    align_dendro(method = "ward.D2") +
    theme_no_axes("x") +
    no_expansion("x") +

    # step into the parent stack
    stack_active() +

    # the original author don't present the full range of the signatures for
    # each hub signature, we just link the single signature instead
    cross_link(
        link_tetragon(
            !!!lapply(subtypes, function(nm) nm ~ nm),
            .element = element_polygon(
                fill = RColorBrewer::brewer.pal(5L, "Set2"),
                color = NA
            )
        ),
        # the data will be used for the layout
        data = t(subtype_scores[, 2:6]),
        inherit_index = TRUE,
        size = 0.1
    ) +
    align_group(factor(
        names(subtype_scores)[-1L],
        c("Wound Healing", "Lymphocyte", "Macrophage", "TGF-beta", "IFN-gamma")
    )) +

    # add second heatmap
    ggheatmap() +
    labs(fill = "scores") +
    scheme_align(NULL) +
    facet_grid(
        labeller = labeller(.panel_y = function(x) {
            dplyr::if_else(x == "C5", "C5  \n", x)
        })
    ) +
    theme(strip.clip = "off") +

    # split by immune subtypes
    anno_left() +
    align_group(forcats::fct_rev(subtype_scores[["Immune Subtype"]])) +

    # we add bottom annotation
    anno_bottom() -
    scheme_align(free_spaces = "l") +
    ggfree() +
    scheme_align(NULL) +
    scheme_data(function(x) {
        dplyr::mutate(x,
            value = scale(value, center = FALSE),
            scaled_scores = mean(value),
            .by = c(.panel, .extra_panel)
        )
    }) +
    geom_density(aes(value, fill = scaled_scores), color = NA) +
    geom_vline(xintercept = 0, linetype = "dashed") +
    scale_fill_gradient2("scaled\nscores", low = "blue", high = "red") +
    facet_grid(
        rows = vars(.extra_panel),
        cols = vars(.panel), as.table = FALSE,
        axes = "all_x"
    ) +
    theme_bw() +
    theme_no_axes() +
    theme(
        strip.text.x = element_blank(),
        strip.text.y = element_text(size = 14),
        strip.background = element_blank(),
        panel.border = element_blank(),
        axis.line.x.bottom = element_line(),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 14, face = "bold")
    )
ggsave("output/FigureS6.pdf",
    plot = p, width = 7, height = 10,
    family = "Helvetica"
)
```
